---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
data <- read.csv("activity.csv", header = TRUE)
summary(data)
```

## What is mean total number of steps taken per day?
```{r}
total.steps <- tapply(data$steps, data$date, sum, na.rm=TRUE)
hist(total.steps, breaks = 50, main = "Mean Total Number of Steps Taken Per Day", xlab = "No. of Steps")
```

Mean number of steps taken per day:
```{r}
mean(total.steps)
```

Median number of steps taken per day:
```{r}
median(total.steps)
```

## What is the average daily activity pattern?
5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):
```{r}
average.steps <- aggregate(x=list(steps=data$steps), by=list(interval=data$interval),FUN=mean, na.rm=TRUE)
ggplot(data=average.steps, aes(x=interval, y=steps)) +
geom_line() +
xlab("Interval Index") +
ylab("Average Number of steps taken")
```

The following 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps:

```{r}
average.steps[which.max(average.steps$steps), "interval"]
```

## Imputing missing values
Total # of Missing Values:
```{r}
sum(is.na(data))
```

Impute missing values with means of the interval:
```{r}
# Replace each missing value with the mean value of its 5-minute interval
replace.missing <- function(steps, interval) {
    imputed <- NA
    if (!is.na(steps))
        imputed <- c(steps)
    else
        imputed <- (average.steps[average.steps$interval==interval, "steps"])
    return(imputed)
}
imputed.data <- data
imputed.data$steps <- mapply(replace.missing, imputed.data$steps, imputed.data$interval)
```

Histogram of total number of steps taken each day with imputed data:
```{r}
total.steps.imputed <- tapply(imputed.data$steps, imputed.data$date, sum, na.rm=TRUE)
hist(total.steps.imputed, breaks = 50, main = "Mean Total Number of Steps Taken Per Day", xlab = "No. of Steps")
```

Mean number of steps taken per day:
```{r}
mean(total.steps.imputed)
```

Median number of steps taken per day:
```{r}
median(total.steps.imputed)
```

Mean and median values are higher after imputing data.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
#Creating a factor variable to determine if date is weekend or weekday
is.day.weekend <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        return("weekday")
}
imputed.data$date <- as.Date(imputed.data$date)
imputed.data$day <- sapply(imputed.data$date, FUN=is.day.weekend)
```

Time Series Plot showing differences in activity patterns between weekdays and weekends:
```{r, echo = FALSE}
par(mfrow=c(1,1))  
imputed.data <- aggregate(steps ~ interval + day, data=imputed.data, mean)
with(imputed.data, plot(steps ~ interval, type="n", main="Weekday vs. Weekend Activity Patterns"))  
with(imputed.data[imputed.data$day == "weekday",], lines(steps ~ interval, type="l", col="blue"))  
with(imputed.data[imputed.data$day == "weekend",], lines(steps ~ interval, type="l", col="red" ))  

legend("topright", lty=c(1,1), col = c("blue", "red"), legend = c("weekday", "weekend"), seg.len=3)
```